---
title: "DA5020.A2.Hsiao-Yu.Peng"
author: "Hsiao-Yu Peng"
date: "2023-09-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)

```

## R Markdown

### Q1-1 See the attached certificate.

### Q1-2
```{r}
# load msleep dataset
data("msleep")

# look over msleep dataset
str(msleep)

# if any missing value
any(is.na(msleep)) 

```

The dataset "msleep" has dimension of 83 rows and 11 columns. It includes missing values within the dataset. The variables in the dataset encompass both character and numerics data types.


### Q2.
```{r}
# find out the pecentage of each type of vore
vore_count <- msleep %>% 
  group_by(vore) %>% 
  summarize(n=n()) %>% 
  mutate(percentage = n/sum(n) * 100)

print(vore_count)

# Visualization for bar chart
ggplot(vore_count, aes(vore, percentage, fill = vore)) + 
  geom_bar(stat = "identity")


```


### Q3.
```{r}
# Extract data for omnivores. Find its mean sleep_total
msleep %>% 
  filter(vore =='omni') %>% 
  summarise(mean(sleep_total))

```

### Q4
```{r}
# Show a scatterplot between bodywt and brainwt
ggplot(msleep) + geom_point(aes(bodywt, brainwt)) +
  scale_x_log10() +
  scale_y_log10() 

```
The scatterplot indicates that there may be a positive correlation between body with (bodywt) and brain weight (brainwt).



### Q5.
```{r}
# transform bodywt, brainwt
log_bw <- log10(msleep$bodywt)
log_brainwt <- log10(msleep$brainwt)

# Calculate Pearson's coefficient
# ignore any NA values: `use = "complete.obs"`
correlation <- cor(log_bw, log_brainwt, use = "complete.obs") 

# result
correlation

```

Yes, the Pearson coefficient of correlation is 0.965, which supports the assumption made in question 4.


### Q6.
```{r}
# Calculate mean and standard deviation of sleep_total
mean_sleep_total <- mean(msleep$sleep_total)
sd_sleep_total <- sd(msleep$sleep_total)

# Identify outliers
outliers <- msleep %>%
  filter(abs(sleep_total - mean_sleep_total) > 1.5 * sd_sleep_total)

# Display the name and sleep_total of outliers
select(outliers, name, sleep_total)

```


