---
title: "DA5020.A1.Hsiao-Yu.Peng"
author: "Hsiao-Yu Peng"
date: "2023-09-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

### Q1-1. 
Please see the attached certificate.

### Q1-2. 
Data frame stores data in a two-dimensional table of columns (variables) and rows (observation). We use it to solve a specific theme. Here is an example:

```{r}
# Data frame example
df <- data.frame(
  patientID = c("U101", "U102", "U103", "U104"),
  gender = c("M", "F", "F", "M"),
  age = c(50, 66, 35, 56),
  grade = c("high", "low", "low", "medium")
  )

df
```


### Q2.

```{r}
cars <- c("Truck", "Car", "SUV")
mpg <- c(11, 30, 24)
cost <- c(45000, 25000, 35000)
DF <- data.frame(cars, mpg, cost)
print(DF)

```

### Q2a.
```{r}

# Select row 1 in column 3, what was selected?
DF[1, 3]

# Select rows 1 through 3, what was selected?
DF[1:3, ]

# Select the last column, what was selected?
DF[, 3]

```

### Q3.
```{r}
# load dataset mtcars
data(mtcars)

# first 3 rows of the dataset
head(mtcars, 3)

# last 5 rows of the dataset
tail(mtcars, 5)

```

### Q4.
```{r}
str(mtcars)

# Categorical variables
cat_var <- c("vs", "am")
cat("Categorical variables: ", cat_var)



# Continuous variables
con_var <- c("mpg", "wt")
cat("Continuous variables: ", con_var)
```

### Q5.
```{r}
library(ggplot2)
data("mtcars")
ggplot(mtcars, aes(x = disp, y=mpg))

```

The code above did not actually displayed a chart, but it creates the base for a scatter plot. To show a scatter plot represents the relationship between "disp" and "mpg", we need to add a geom layer. The updated code is as follows:

```{r}
ggplot(mtcars, aes(x = disp, y=mpg)) + geom_point()

```
### Q6.
```{r}
# load the data frame
data("mpg")

# create a scatter plot: displ vs. hwy
ggplot(data=mpg, aes(displ, hwy, color = cyl)) + geom_point()

```

### Q6 explanation
The scatter plot illustrate the relationship between engine displacement and highway miles per gallon. Each point's color indicates the number of cylinders in the car.  It is evident that fewer engine displacement, for example, an engine displacement of 2, results in higher highway miles per gallon. However, as engine displacement increases, highway miles per gallon decreases. For instance, with an engine displacement of 5 or 6, the car achieves only about 20 miles per gallon or less on the highway.


### Q7.
```{r}
ggplot(data = mpg, aes(displ, hwy, color = cyl)) + geom_point() + facet_wrap(vars(drv))

```


### Q7 explanation
Compared to 4WD and front-wheel drive (FWD), the rear-wheel drive (RWD) generally features engine displacements from 4 to 7. In contrast, FWD typically has engine displacements ranging from 1 to 4, while 4WD varies across several types, with engine displacements ranging from 1 to 7.

In terms of highway miles per gallon (MPG), RWD vehicles generally achieve a range of 15 to 25 MPG. Conversely, FWD vehicles typically achieve 25 to 30 MPG or even more on the highway. As for 4WD vehicles, they exhibit a wide range of MPG values due to various engine displacements. Generally, the higher the engine displacement, the lower the MPG 4WD achieves.

