---
title: "DA5020.A4.Hsiao-Yu.Peng"
author: "Hsiao-Yu Peng"
date: "2023-10-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
```

### Bonus 1. See the attached DataCamp certificates.


### Bonus 2.
```{r, warning==FALSE}
# load dataset
tripdata_df <- read_csv("~/Desktop/2023Fall_Syllabus/DA5020/week3/2018_Green_Taxi_Trip_Data.csv")

```

```{r}
# Add a "pickup" variable which is from "lpep_pickup_datetime" varialbe, convert "picup" data type to Date type.
tripdata_df <- tripdata_df %>% 
  mutate(pickup = as.Date(lpep_pickup_datetime, format = "%m/%d/%Y"))


tripdata_df %>% 
  group_by(pickup) %>% 
  summarize(count = n()) %>% 
  arrange(desc(count))
```

2018-02-02, recorded the highest number of trips, with a total of 33,953 trips on that day. Upon further investigation of date in history, it shows that Fridays and Saturdays have significantly more trips than other weekdays.

### Q1. 
```{r}
glimpse(tripdata_df)

# Check documenatation. VendorID, ratecodeID, payment_type, trip_type are factor variables, we can covert them to a factor.
variables_cat <- c("VendorID", "RatecodeID", "payment_type", "trip_type")

# Use lapply to convert the categorical variables
tripdata_df[variables_cat] <- lapply(tripdata_df[variables_cat], as.factor)

# Check the levels of each variable's factor
for (var in variables_cat) {
  cat(sprintf("%s factor is: %s\n", var, paste(levels(tripdata_df[[var]]), collapse = ", ")))
}
```
The dataset has dimension 1,048,575 x 20. The data type of "lpep_pickup_datetime" and "lpep_dropoff_datetime" are character, not datetime. Other categorical variables like VendorID, RatecodeID, trip_type, and payment_type have "numerical" datatype. So we convert them in factor type here.

It's worth noting that the RatecodeID factor includes the value '99,' which is considered invalid in the dataset. 


### Q2
```{r}
# Analyze Trip Types (Hailing Method)
trip_type_analysis <- tripdata_df %>%
  group_by(trip_type) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) 

# Analyze Payment Types
payment_type_analysis <- tripdata_df %>%
  group_by(payment_type) %>%
  summarize(count = n()) %>%
  arrange(desc(count)) 

trip_type_analysis
payment_type_analysis

# Analyze what payment type is the most when people hail a cab
payment_type_hailcab <- tripdata_df %>%
  filter(trip_type == 1) %>%
  group_by(payment_type) %>%
  summarize(counts = n())

payment_type_hailcab
  
```
Trip type 1, which is street-hail, is the most common way people hail a cab, with a total count of 1,030,020 trips.

Payment type 1, which is credit card, is the most common way people pay for their trips, with a total count of 564,347.

In further analysis, we found Payment_type 1 (credit_card) is the most common method when people use "street-hail", with a total count of 557,317.



### Q3
```{r}
# February pickups
feb_pickups <- tripdata_df %>%
  filter(pickup >= "2018-02-01" & pickup < "2018-03-01") %>%
  group_by(pickup) %>% 
  summarize(count = n())

# barplot
ggplot(feb_pickups, aes(x = pickup, y = count)) + 
  geom_bar(stat = "identity") +
  labs(x = "Date (February)", y = "Frequency of Pickups") + 
  ggtitle("Frequency of Picups in February")
```


From the bar plot, we observed that the majority of pickups occurred during the early part of February, specifically from February 1st to February 9th. After February 10th, the pickup frequency dramatically decreased to single digits, such as 3 or 6, and remained low.


### Q4
```{r}

HourOfDay <- function(timestamp){

  # Use regular expression to extract the hour (HH) part
  hour <- sub(".*\\s(\\d{1,2}):.*", "\\1", timestamp)
  return(hour)
}

# Example
timestamp <- "9/23/2010 11:17"
hour <- HourOfDay(timestamp)
cat("Hour:", hour, "\n")

```


### Q5
```{r}

# Extract Hour element in all Pickup date
tripdata_df <- tripdata_df %>% 
  mutate(lpep_pickup_hour = HourOfDay(lpep_pickup_datetime))

tripdata_df %>% 
  select(lpep_pickup_hour) %>% 
  head()

```


### Q6
```{r}
median_distance <- tripdata_df %>% 
  group_by(lpep_pickup_hour) %>% 
  summarize(median_dist = median(trip_distance)) 
 

# Convert "lpep_pickup_hour" to numeric type
median_distance$lpep_pickup_hour <- as.numeric(median_distance$lpep_pickup_hour)

# Create the bar plot
ggplot(median_distance, aes(x = lpep_pickup_hour, y = median_dist)) +
  geom_bar(stat = "identity") +
  labs(x = "Hour of Day", y = "Median Trip Distance") +
  ggtitle("Median Trip Distance by Hour of Day")

```

The longest median trip distance occurs during pickups at 5 AM and 6 AM in the morning. It then gradually decreases as the day progresses, reaching its shortest median trip distance at 19:00. After 20:00, the median distance begins to gradullay increase once again.
