---
title: "DA5020.A6.Hsiao-Yu.Peng"
author: "Hsiao-Yu Peng"
date: "2023-10-22"
output: pdf_document
---

# Bonus 1: See the attachement.


# Bonus 2: What is the average runtime for the thriller movie genre.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(RSQLite)
library(ggplot2)
```

```{r}
db_path <- "~/Downloads/sqlite-tools-osx-x86-3430200/imdb.db"

# Connect to the database
con <- dbConnect(SQLite(), dbname = db_path)

# Define the query
query <- "SELECT Genre, AVG(Runtime) AS Avg_Runtime
          FROM movie_info
          WHERE Genre = 'Thriller';"

# Execute the query and fetch the result
result <- dbGetQuery(con, query)

# Close the database connection
dbDisconnect(con)

# print the result
print(result)
```
The average runtime for the thriller movie genre is 108 minutes.


# Q1
# Q1-1 sql create the table director_info   
#CREATE TABLE director_info (  
#    Director_ID INT AUTO_INCREMENT PRIMARY KEY,  
#    Director_Name VARCHAR(255) NOT NULL  
#);  


# Q1-2 sql import csv to director_info
#.mode csv  
#.import --skip 1 directors.csv director_info

# - Verify the imported data  
#SELECT * FROM director_info;


# Q2
# Q2-1. Count the number of rows in the movie_info and director_info tables.
```{r}
db_path <- "~/Downloads/sqlite-tools-osx-x86-3430200/imdb.db"
# Connect to the database
con <- dbConnect(SQLite(), dbname = db_path)

# query
query_movie_count <- "Select COUNT(*) from movie_info;"
query_director_count <- "Select Count(*) from director_info;"

# fetch the results
movie_count <- dbGetQuery(con, query_movie_count)
director_count <- dbGetQuery(con, query_director_count)

# Close the database connection
dbDisconnect(con)


# Print out the counts
cat("Number of rows in movie_info:", movie_count[1, 1], "\n")
cat("Number of rows in director_info:", director_count[1, 1], "\n")
```


# Q2-2. How many movies were released between 2010 and 2020 (inclusive)? Visualize the results.

```{r}
# Define the path to the SQLite database
db_path <- "~/Downloads/sqlite-tools-osx-x86-3430200/imdb.db" 

# Connect to the database
con <- dbConnect(SQLite(), dbname = db_path)

# Define the SQL query 
query <- "SELECT COUNT(*) as Count, Release_Year as Year
          FROM movie_info
          WHERE CAST(Release_Year AS INTEGER) BETWEEN 2010 AND 2020
          GROUP BY Year;"

# Execute the query and fetch the results
result <- dbGetQuery(con, query)
print(result)

# Visualize the results
ggplot(result, aes(x = Year, y = Count, fill = Year)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Movies Released between 2010 and 2020",
       x = "Year", y = "Count") +
  theme_minimal()

# Close the database connection
dbDisconnect(con)
```


# Q2-3 What is the minimum, average and maximum ratings for “Action” movies. Ensure that you query the genre using wild cards.
```{r}
# Define the path to the SQLite database
db_path <- "~/Downloads/sqlite-tools-osx-x86-3430200/imdb.db" 

# Connect to the database
con <- dbConnect(SQLite(), dbname = db_path)

# Define the SQL query
query <- "SELECT MIN(IMDB_Rating),
                 AVG(IMDB_Rating),
                 MAX(IMDB_Rating)
          FROM movie_info
          WHERE Genre LIKE '%Action%';"

# Execute the query and fetch the results
result <- dbGetQuery(con, query)

# Print the results
print(result)

# Close the database connection
dbDisconnect(con)
```
For Action movies, the minimum rating is 7.6, the average rating is 7.94, and the maximum rating is 9.


# Q2-4 What are the 25 highest-grossing movies within the dataset? Display the title, genre and gross.

```{r}
# Define the path to the SQLite database
db_path <- "~/Downloads/sqlite-tools-osx-x86-3430200/imdb.db" 

# Connect to the database
con <- dbConnect(SQLite(), dbname = db_path)

# Define the SQL query
query <- "SELECT Series_Title, Genre, CAST(Gross AS NUMERIC) AS Gross
          FROM movie_info
          WHERE Gross IS NOT NULL
          ORDER BY Gross DESC
          LIMIT 25;"

# Execute the query and fetch the results
result <- dbGetQuery(con, query)

# Print the results
print(result)

# Close the database connection
dbDisconnect(con)

```


# Q2-5. Which directors have the highest-grossing movies. Display the director name and the total gross. Ensure that you join the necessary tables. Visualize the results using a Bar chart.

```{r}
# Define the path to the SQLite database
db_path <- "~/Downloads/sqlite-tools-osx-x86-3430200/imdb.db" 

# Connect to the database
con <- dbConnect(SQLite(), dbname = db_path)

query <- "SELECT d.Director_Name, SUM(CAST(Gross AS NUMERIC)) AS Total_Gross
          FROM director_info d 
          JOIN movie_info m ON d.Director_ID = m.Director_ID
          WHERE m.Gross IS NOT NULL
          GROUP BY Director_Name
          ORDER BY Total_Gross DESC
          LIMIT 5;"

# Execute the query and fetch the results
result <- dbGetQuery(con, query)

# Close the database connection
dbDisconnect(con)

# Visualize the results with a bar chart
ggplot(result, aes(x = Director_Name, y = Total_Gross)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Gross for Top-5 Directors with Highest-Grossing Movies",
       x = "Director Name", y = "Total Gross") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Steven Spielberg has the highest-grossing movies.


# Q2-6 Create a function called verifyDirector() that takes a director name as its argument, and queries the database to check if the director exists. Your function should display a message to notify the user if the director was found or not.

```{r}
verifyDirector <- function(director_name) {
  # Define the path to the SQLite database
  db_path <- "~/Downloads/sqlite-tools-osx-x86-3430200/imdb.db" 
  
  # Connect to the database
  con <- dbConnect(SQLite(), dbname = db_path)
  
  # Define the SQL query 
  query <- sprintf("SELECT 1
                    FROM director_info
                    WHERE Director_Name = '%s'
                    LIMIT 1;", director_name)
  
  # Execute the query and fetch the result
  result <- dbGetQuery(con, query)

  # Close the database connection
  dbDisconnect(con)
  
  # Check the result and display a message
  if (nrow(result) > 0) {
    cat(sprintf("Director '%s' found in the database.\n", director_name))
  } else {
    cat(sprintf("Director '%s' not found in the database.\n", director_name))
  }
}

# Example for the function
verifyDirector("Steven Spielberg")
verifyDirector("Ben Tasker")
```